---
title: "R Tutorial"
output: html_notebook
---

# Install and load packages
```{r}
# # installing packages
# # people make and upload packages to CRAN; you can download more from there
# install.packages("tidyverse") # "equivalent of Pandas on steroids" - Kajal 
# install.packages("ggplot2")
# 
# # loading the packages into R runtime environment
# library(tidyverse) 
# library(ggplot2)

# this first block of code should only be run once, which is why I commented it out. You don't need it in your code, as the second block of code does it for you. 
```

```{r}
# this block of code only installs the packages if they aren't already installed
.packages <- c("tidyverse", "ggplot2") # these are the packages in interest
# c("tidyverse","ggplot2") makes a vector/list of the databases
# Install CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])
# Load packages into session
lapply(.packages, library, character.only=TRUE)
```

We are going to duplicate the Pandas tutorial in R!

Dataset for tutorial and problem set:
Pima Indian Heritage Diabetes

Each person in the dataset is a female who is at least 21 years old

Below are the variables for each person in the dataset
  'preg': number of pregnancies  
  'plas': plasma glucose concentration 
  'pres': blood pressure 
  'skin': skin thickness
  'test': Insulin
  'mass': BMI
  'pedi': diabetes pedigree function
  'age': age
  'class': '0' means does not have diabetes and '1' means has diabetes
Go to the link below to see the raw comma seperated values (csv) data:

https://raw.githubusercontent.com/jbrownlee/Datasets/master/pima-indians-diabetes.data.csv

Step 1: Read the data source into a dataframe
```{r}
## Read the csv file
df <- read_csv("https://raw.githubusercontent.com/jbrownlee/Datasets/master/pima-indians-diabetes.data.csv", col_names = c("preg", "plas", "pres", "skin", "test", "mass", "predi", "age", "class"))

```

Look at the data frame
```{r}
## the type 
class(df)
```
Examine the data frame
```{r}
# number of rows
nrow(df) # run on any dataframe to get # of rows in the dataframe
sprintf("Number of rows:%d", nrow(df)) # easy way of printing things out

# number of columns
ncol (df)
sprintf("Number of columns:%d", ncol(df))

## get the column names
colnames(df)

# prints top x number of rows
head(df) ## 6 rows is the default
head(df, 10) ## gives you the top 10 rows 
```
Some other functions for looking at the data
```{r}
glimpse(df) ## from tidyverse
# gives you a glimpse of the entire dataframe
# tells you num rows, num variables, give some of the values
# will take a long time because we have a lot of variables and rows
```

Indexing, filtering, and summarizing data.

There are multiple ways of indexing, filtering and summarizing data. We will look at two flavors of it.

First - Using array methods
```{r}
## get the first row
df[1,] # similar to numpy slicing
# get first row and all of columns

## get the 3rd column for the first 10 rows
df[1:10, 3]


# get columns 1 through 5 for all rows
df[, 1:5]

# you can also use column names to retrieve 
## get the pres column for the first 10 rows
df[1:10, "pres"]

## to assign to a new variable
a <- df [,1:5]
# <- is how you assign vars in R

head(a)

```

Living in the tidyverse - and filtering. 
"%>%" -- powerful chaining/pipelining method
```{r}
## get the first row only
df %>% # %>% => takes output of first method and passes it as input to the next
  slice(1) 

## get the 3rd column of the first 10 rows (using column position)
df %>%
  slice (1:10) %>%
  select (3) # selects column 3 of the dataframe made in the line above

# get columns 1 through 5 for all rows
df %>%
  select (1:5) ############################### SELECT METHOD #################################

# using column names (the more widely used method)
## get the 3rd (pres) column for the first 10 rows
df %>%
  slice (1:10) %>% 
  select (pres) # select the pres column of the output of the line above

# multiple columns
df %>%
  slice (1:10) %>%
  select (preg, pres, class)

# assignment works the same way
a <- df %>%
     slice (1:10) %>%
     select (preg, pres, class)

############################################## NOTE: To save any of the data, you must assign them to a variable!!
```

Lets do some filtering!
In this example, we're going to subset into this dataset to get only women over 39 who have had more than 2 pregnancies.

```{r}
df %>%
  filter(age > 39 & preg > 2) 

## you can assign this to another data frame
mydf <- df %>%
        filter(age > 39 & preg > 2)

head(mydf)
```

Summary statistics on the data
```{r}
## Find the mean of a column
mean(df$preg) # mean number of pregnancies
# '$' gets the variable preg in df

sd (df$preg) # standard deviation
quantile (df$preg)

## doing it across mulitple columns
df %>%
  summarize(mean_preg = mean (preg),
            mean_plas = mean (plas),
            mean_pres = mean (pres))

```

Finding and dealing with missing values

```{r}
mydf %>%
  select (everything()) %>%
  summarise_all(funs(sum(is.na(.))))
```

Missing data - find it and deal with it!

Download the small subset of NHANES from: https://drive.google.com/open?id=1hW0wgPc-tIiHbHCHMLWUvVV_xv2km1TH
```{r}
nh_df <- readRDS ("~/Downloads/small_subset.rds")

nrow(nh_df)
ncol(nh_df)
```

How much data is missing? R marks missing data as NAs
is.na () -- powerful function for finding the NAs
```{r}
nh_df %>%
  select (everything()) %>%
  summarise_all(funs(sum(is.na(.))))

```


Summary statistics with NAs
```{r}
## Find the mean of a column
mean(nh_df$BMXBMI, na.rm=TRUE) # mean number of pregnancies
sd (nh_df$BMXBMI, na.rm=TRUE) # standard deviation
quantile (nh_df$BMXBMI, na.rm=TRUE)

## doing it across mulitple columns
df %>%
  summarize(mean_bmi = mean (BMXBMI, na.rm =TRUE),
            mean_age = mean (RIDAGEYR, na.rm=TRUE))
```

